<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Introduction to qtlpvl}
-->

```{r, echo = FALSE, message = FALSE}
library(qtlpvl)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```

# Introduction to qtlpvl

This package is focusing on QTL mapping with multiple traits and
testing of pleiotrophy vs. close linkage.


## Data

We use data `listeria` from `R\qtl` as an example. We will be only
using the genotypes of this F$_2$ population and generate the
phenotypes from normal distribution.

```{r}
set.seed(92950640)
data(listeria)
listeria <- calc.genoprob(listeria,step=1)
n <- nind(listeria)
chr <- "1"
geno <- pull.geno(listeria, chr=chr)
genotype1 <- geno[,7]
genotype2 <- geno[,10]
(pos <- unlist(pull.map(listeria, chr=chr))[c(7,10)])
p <- 10
p1 <- floor(p/2)
G1 <- matrix(genotype1, n, p1)
G2 <- matrix(genotype2, n, p-p1)
G2[G2==3] <- 2
G <- cbind(G1, G2*(-2))
Y <- matrix(rnorm(n*p),n,p)
Y <- Y + G
```

In the above code, after setting the random seed, we load the data
`listeria` and pull out the genotype data from the cross object
`listeria` on chromosome `r chr` as a data matrix `geno`, which is of
size `r n` $\times$ `r ncol(geno)`. We pick the 7th and 10th marker as
the two QTLs and save their genotypes in `genotype1` and
`genotype2`. These two QTLs are at position `r sprintf("%.2f",pos[1])`
cM and `r sprintf("%.2f",pos[2])` cM on chromosome `r chr`.

Assume the first QTL have additive effect only and the second QTL have
dominant effect --individuals with genotype 2 and 3 are grouped
together to have the same shift in mean.  The first 5 traits are
controlled by the first QTL and the other 5 trait of the total 10 are
controlled by the second QTL with negative effect.  We use matrix G to
save the QTL effects and the second QTL group, `G2` is multiply by -2
to have a negative QTL effect.  The 10 phenotypes are generated with
these QTL effect matrix G plus independent noise generated from
standard normal distribution.

We could first look at the single triat analysis result.  For each
trait, we plot its LOD score versus QTL position when the LOD is
bigger than a threshold (default value is 3).

```{r}
plotLOD(listeria, Y, chr)
```


## joint mapping: `scanone.mvn`

Having the cross object `listeria` and phenotype trait matrix `Y`
prepared, we can first run a joint mapping with `scanone.mvn`, which
assumes multivariate normal model. chr is set to be "1", which
indicates the mapping is only on chromosome 1. Function `scanone.mvn`
has two more parameters `addcovar` and `intcovar`, they could be used
to control for additive and interactive covariates when needed.

The result of joint mapping is a data frame with class `scanone`, we
could apply (S3) method `summary` and `plot` to look at the result.

```{r}
out <- scanone.mvn(listeria, Y, chr)
summary(out)
plot(out)
```

We could see the QTL is at `r sprintf("%.2f",summary(out)$pos)` cM
with LOD score `r sprintf("%.2f",summary(out)$lod)`.

## test of 1 vs 2 QTL: `testpleio.1vs2`

We test the hypothesis of: 
* $H_0$: There is only one QTL affecting all the phenotypes;
* $H_1$: There are two QTLs, each affecting one of the two phenotype
  groups.

We assume the traits follow multivariate normal distribution and
perform joint mapping to locate the single-QTL under the null
hypothesis that there is one pleiotrophic QTL affecting all the
traits; we then run single trait analysis on each trait and find the
trait-specific QTL, sort the traits by their QTL position and search
for the best separation of the traits into two groups, where the first
group (contains the first several traits) is controlled by the right
QTL and the second group (contains the rest traits) is controlled by
the left QTL. The LOD score of this two-QTL model is then subtracted
by the LOD score of the single-QTL model to arrived at the final test
statistic.

To get the null distribution of the statistics, we have two methods:
* parametric bootstrap: generate data from the parameters estimated
  under the null hypothesis
* stratified permutation: randomly permute the genotype data versus
  phenotype data with in each QTL genotype group,

we repeat the whole procedure on data from either method and save the
test statistics. The P-value is calculated from this empirical
distribution.

Function `testpleio.1vs2` is used to do this test. input parameters:
* cross: An object of class `cross`. 
* Y: matrix of multiple traits with n rows.
* chr: One character strings referring to chromosome of interest by
  name.
* region.l, region.r: left and right bound for the interval of
  interest.
* int.method: can have value of `bayes` or `1.5lod`, method to
  calculated the interval of interest if `region.l` and `region.r` is
  not specified.
* search.method: searching method for two-QTL model, can have value of
  `fast` or `complete`, default is `fast`.
* RandomStart: use random starting point for the two-QTL model or not,
  default is `TRUE`.
* RandomCut: use random cutting or not when there are traits mapped to
  the same location. default is `FALSE`.
* simu.method: `parametric` or `permutation` method for
  simulations. default is `parametric`.
* n.simu: number of simulations for p-value.
* tol: Tolerance value for the `qr` decomposition in `lm` fitting.
* addcovar, intcovar: Additive and interactive covariates to control
  for when mapping.

This function will return a list with class `testpleio.1vs2`, similar
to `scanone`, we could apply (S3) method `summary` and `plot` to look
at the result. In the summary, we could see the position and LOD score
for the single QTL and best two QTL model, as well as test statistics
LODdiff=LOD2-LOD1 and its P-value.

```{r}
obj <- testpleio.1vs2(listeria, Y, chr, n.simu=100, region.l=60, region.r=90)
summary(obj)
plot(obj)
```

The above figure shows the joint mapping result in black and profile
for each of the two QTLs under $H_1$:

* red line: $L_1(\lambda_1) = \max_{\lambda_2} LOD2(\lambda_1,
  \lambda_2)$
* blue line: $L_2(\lambda_2) = \max_{\lambda_1} LOD2(\lambda_1,
  \lambda_2)$

Triangles indicate the position of the QTLs and solid points indicate
QTL positions and LODs from single trait mapping.

We could also use `plottrace` to see how LODdiff changes when we move
the cut-point of the left versus right group. There are 7 possible QTL
positions from the above plot (result of single trait mapping), thus
we have 6 possible ways of grouping these positions into the left
versus right group. The 6 open dot indict the cut and their LODdiff.
Thus when the first 5 traits and the last 5 traits are grouped
separately, the LODdiff is the biggest, it is our best two QTL model.
This figure could be used as a diagnostic figure, to see if our best
two QTL model is better than the rest of all the possible models.

```{r}
plottrace(obj)
```

## test of 1 vs p QTL: `testpleio.1vsp`

We could also test the hypothesis of:
* $H_0$: There is only one QTL affecting all the phenotypes; 
* $H_1$: There are p QTLs, each affecting one of the phenotypes, i.e.,
  each phenotype is allowed to has its own QTL.

This function has much less parameters than the previous one. We have
`cross`, `Y`, `chr`, `addcovar`, `intcovar` and `n.simu`, the usage is
same as before.  This function will return a list with class
`testpleio.1vsp`, similar to `scanone` and `testpleio.1vs2`, we could
apply (S3) method `summary` and `plot` to look at the result.  In the
summary, we could see the position and LOD score for the single QTL
and p QTL model, as well as test statistics LODdiff=LODp-LOD1 and its
P-value.

```{r}
obj2 <- testpleio.1vsp(listeria, Y, chr, n.simu=100)
summary(obj2)
plot(obj2)
```

## plot Genetic Pattern: `plotGenetpattern`

To illustrate the use of exploration plots, we generate data with more
number of traits, 100. Same as before, the first half of the traits
are controlled by the first QTL and the rest are controlled by the
second QTL.

```{r}
p <- 100
p1 <- floor(p/2)
G1 <- matrix(genotype1, n, p1)
G2 <- matrix(genotype2, n, p-p1)
G2[G2==3] <- 2
G <- cbind(G1, G2*(-2))
Y <- matrix(rnorm(n*p),n,p)
Y <- Y + G
```

Function `plotGenetpattern` takes two kinds of input parameters:
* a phenotype matrix `Y` and a genotype vector `genotype`, the later
  should be genotype of the common QTL in most suspicious.
* a phenotype matrix `Y` and a cross object and chromosome
  number. This is used when there is no certainty of a single QTL for
  all traits and thus each trait is mapped separately and its own QTL
  genotype is used.
  
QTL additive effect is defined as a=(RR-BB)/2, QTL dominant effect is
defined as d=BR-(BB+RR)/2, when plotted against each other, traits
with pure additive effects are near the x-axis and traits with
dominant effects are along the diagonals, $y=x$ and $y=-x$.

```{r}
plotGenetpattern(Y, genotype=genotype1)
plotGenetpattern(Y, cross=listeria, chr=chr)
```

We can tell from both plots the there are two different genetic
patterns, thus possible two different QTLs, even if they are closely
linked.

## plot LOD score with sign: `plotLODsign` 

Using the same data, we could also display the direction of the QTL
effect with `plotLODsign`.  Function `plotLODsign` takes input
`cross`, `Y` and `chr`, as well as `addcovar` and `intcovar`, all has
been introduced above.

For each trait, the QTL effect direction is used as the sign of the
LOD score. We first run single trait mapping for LOD scores and QTL
positions, then we use `argmax.geno` to impute QTL genotypes and
calculate mean for each of the 3 genotype group. We define the
direction positive when the mean of the first group is bigger than the
mean of the third group, and negative while not.

```{r}
plotLODsign(listeria, Y, chr)
```

## Session Info
```{r}
sessionInfo()
```
