<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Introduction to qtlpvl}
-->

```{r, echo = FALSE, message = FALSE}
set.seed(20150805)
knitr::opts_chunk$set(
    fig.width=10, fig.height=6,
    comment = "#>",
    error = FALSE,
    tidy = FALSE)
```

# Introduction to qtlpvl

This package focuses on QTL mapping with multiple traits and
on testing of pleiotrophy vs. close linkage when multiple traits map
near each other, particularly for the of a trans-eQTL hotspot.
We provide both exploratory plots and formal
statistical tests aimed at dissecting trans-eQTL hotspots.

## Data

To illustrate the package, we use the genotype data from the
`listeria` data set (an F$_2$ population) provided with [R/qtl](http://rqtl.org)
along with a set of simulated phenotypes, included in qtlpvl as the
data set `fake.phenos`. The phenotypes were simulated using
two markers from chromosome 1 as QTL,
with the first QTL having an additive allelic effect, and with one of
the alleles at the second QTL being strictly dominant. There are 10
phenotypes. The first 5
are controlled by the first QTL, and the other 5 traits are controlled
by the second QTL (and with a negative and larger effect).
The 10 phenotypes were generated with these QTL effects plus
independent, normally distributed residual variation.
Treating these traits as gene expression measurements, we
assigned genomic positions at random. The phenotype data is stored
in matrix `Y` and their positions are stored in data frame `probepos`.

We load the qtlpvl package, the simulated phenotype data, and the
`listeria` data set as follows. (The R/qtl package is automatically
loaded when qtlpvl is loaded.)

```{r}
library(qtlpvl)
data(fake.phenos)
data(listeria)
```

We will first perform single-trait analysis. For each
trait, we plot its LOD score versus its QTL position, when the LOD is
bigger than a threshold (the default value is 3). To do so we first
calculate the QTL genotype probabilities for `listeria` and then use
the function `plotLOD`. We will only scan chromosome 1, and we set the
variable `chr=1` for use throughout.

```{r,plotLOD,fig.cap="LOD score and QTL positions for each of the simulated traits"}
listeria <- calc.genoprob(listeria, step=1)
chr <- 1
Y <- fake.phenos
plotLOD(Y=Y, cross=listeria, chr=chr)
```

## joint mapping: `scanone.mvn`

Having the cross object `listeria` and phenotype trait matrix `Y`
loaded, we can perform joint, multi-trait QTL mapping with the
function `scanone.mvn`. This assumes
a multivariate normal model. `chr` is set to be "1", which
indicates the mapping is only on chromosome 1. Function `scanone.mvn`
has two more parameters `addcovar` and `intcovar` that can be used
to control for additive and interactive covariates.
`intcovar` will also be used as additive covariates during
mapping, and there is no need to manually add them into `addcovar`.

The result of joint mapping is a data frame with class `scanone` (that
is, in the same form as returned by the R/qtl function `scanone`).
We can then use `summary` and `plot` to look at the results.

```{r,scanonemvn, fig.cap="Joint mapping result under multivariate normal model for all the simulated triats"}
out <- scanone.mvn(cross=listeria, Y=Y, chr=chr)
summary(out)
plot(out)
```

With this single-QTL analysis, the QTL is estimated to be at
`r broman::myround(summary(out)$pos, 1)` cM,
with LOD score `r broman::myround(summary(out)$lod, 1)`. This is closer to the
second QTL, as in our simulation, the second QTL had a larger effect than
the first QTL.

## Statistical Testing
### test of 1 vs 2 QTL: `testpleio.1vs2`

We now turn to the question of pleiotropy vs. close
linkage. Specifically, we test the hypotheses:
* $H_0$: There is only one QTL affecting all the phenotypes;
* $H_1$: There are two QTL, with each phenotype affected by one or
  the other QTL.

We again assume that the residual variation in the traits follows
a multivariate normal distribution. We
perform joint, multivariate QTL mapping to locate the single QTL, under the null
hypothesis that there is one pleiotropic QTL affecting all the
traits. We then run single-trait analysis on each trait and find the
trait-specific QTL, sort the traits by their estimated QTL position and search
for the best separation of the traits into two groups, where the first
group (contains the first several traits) is controlled by the right
QTL and the second group (contains the rest traits) is controlled by
the left QTL. The LOD score for this two-QTL model is then subtracted
by the LOD score for the single-QTL model, to give the final test
statistic, $\text{LOD}_{2v1}$.

To get the null distribution of the statistics, we have two methods:
* parametric bootstrap: generate data from the parameters estimated
  under the null hypothesis
* stratified permutation: randomly permute the genotype data versus
  phenotype data within each QTL genotype group,

We repeat the entire procedure on data from either method and save the
test statistics. The P-value is calculated from this empirical
distribution.

The function `testpleio.1vs2` is used to do this test. Input parameters:
* `cross`: An [R/qtl](http://rqtl.org)  `cross` object.
* `Y`: matrix of multiple traits, with samples in the row, traits in the
  column.
* `chr`: Character string referring to chromosome of interest by
  name.
* `region.l`, `region.r`: left and right bounds for the interval of
  interest.
* `int.method`: method to
  calculate the interval of interest if `region.l` and `region.r` is
  not specified (either `bayes` or `1.5lod`)
* `search.method`: method for searching the two-QTL model (either
  `fast` or `complete`; the default is `fast`).
* `RandomStart`: When `search.method` is `fast`, indicates whether to use a random starting point
  for the search over two-QTL models; the default is `TRUE`.
* `RandomCut`: Indicates whether to use random cutting or not when there are traits mapped to
  the same location. The default is `FALSE` (traits mapped to the same
  location will be bound together).
* `simu.method`: method for determining the null distribution
  (`parametric` or `permutation`; the default is `parametric`).
* `n.simu`: number of simulations for p-value.
* `tol`: Tolerance value for the `qr` decomposition in fitting the
  linear models with `lm`.
* `addcovar`, `intcovar`: Optional additive and interactive covariates to include
  when mapping.

This function will return a list with class `testpleio.1vs2`. We can
use `summary` and `plot` to study the results.
The result of `summary` includes the estimated QTL position and LOD score
for the single-QTL and two-QTL models, as well as the test statistic
$\text{LOD}_{2v1}$ and its P-value.

As a quick illustration, we will perform just 10 bootstrap simulation replicates.

```{r, fig.cap="LOD profile plot for result of test1vs2 on simulated traits"}
obj <- testpleio.1vs2(cross=listeria, Y=Y, chr=chr, n.simu=10,
                      region.l=60, region.r=90,
                      search.method="complete")
summary(obj)
plot(obj)
```

The above figure shows the joint mapping result in black and the
profile LOD curves for each of the two QTL under $H_1$:

* red line: $L_1(\lambda_1) = \max_{\lambda_2} LOD2(\lambda_1,
  \lambda_2)$
* blue line: $L_2(\lambda_2) = \max_{\lambda_1} LOD2(\lambda_1,
  \lambda_2)$

Triangles indicate the estimated positions of the QTL and solid points indicate
QTL positions and LODs from single trait mapping.

We use `plottrace` to see how $\text{LOD}_{2v1}$ changes
when we move the cut-point of the left versus right group.

```{r, fig.cap="plot of LOD$_{2v1}$ by cut-point"}
plottrace(obj)
```

There are 7
possible QTL positions from the above plot with the results of single-trait
mapping. Thus we have 6 possible ways of grouping these positions
into the left versus right group. The 6 dots indicate the possible cut-points and
the corresponding value for  $\text{LOD}_{2v1}$.Thus when the first 5 traits and the last 5
traits are grouped separately, the $\text{LOD}_{2v1}$ is the biggest,
it is our best two QTL model. This figure can be used as a
diagnostic, to see how our inferred two-QTL model compares to other
possible models.

### test of 1 vs p QTL: `testpleio.1vsp`

We could also test the hypotheses:
* $H_0$: There is only one QTL affecting all the phenotypes;
* $H_1$: There are p QTL, each affecting one of the phenotypes (i.e.,
  each phenotype is allowed to have its own QTL).

This function has fewer parameters than `testpleio.1vs2`. The
arguments are `cross`, `Y`, `chr`, `addcovar`, `intcovar` and
`n.simu`, and their usage is
the same as before.

This function will return a list with class
`testpleio.1vsp`. We can again use `summary` and `plot` to look at the results. The
summary includes the estimated QTL positions and LOD score for the single-QTL
and p-QTL models, as well as test statistics $\text{LOD}_{pv1}$ and its
P-value.

```{r, fig.cap="plot for result of testpleio.1vsp, showing joing mapping LOD curve and max LOD score for each trait seperately"}
obj2 <- testpleio.1vsp(cross=listeria, Y=Y, chr=chr, n.simu=10)
summary(obj2)
plot(obj2)
```

## Exploratry plots

### plot of genetic pattern: `plotGenetpattern`

The function `plotGenetpattern` takes two kinds of input parameters:
* a phenotype matrix `Y` and a genotype vector `genotype`, the later
  being genotypes at a common QTL.
* a phenotype matrix `Y` and a cross object `cross` and chromosome
  number `chr`. This is used when there is uncertainty in the QTL
  locations for the traits. In this case, each trait is mapped
  separately and the genotypes at corresponding estimated QTL
  position are used.

We'll use the labels B and R for the alleles in the cross. (These
correspond to the strains in the main application that motivated this
work).
The additive QTL effect is defined as $a=(\mu_{RR}-\mu_{BB})/2$. The dominance effect is
$d=\mu_{BR}-(\mu_{BB}+\mu_{RR})/2$. When plotted against each other, traits
with pure additive effects are near the x-axis ($d=0$) and traits with
dominant effects are along the diagonals, $d=a$ (that is, $\mu_{BR}=\mu_{RR}$) and $d=-a$
(that is, $\mu_{BR}=\mu_{BB}$).

Here, we'll make plots by each method: with a common QTL and with
separate QTL for each trait.

```{r, fig.width=12, fig.height=6, fig.cap="plot of inheritance pattern for all the simulated traits"}
par(mfrow=c(1,2))
qtlpos <- max(out)$pos
m <- find.pseudomarker(listeria, chr, qtlpos, "prob", addchr=FALSE)
qtlgeno <- apply(listeria$geno[[chr]]$prob[,m,], 1, which.max)

plotGenetpattern(Y=Y, genotype=qtlgeno, main="by common QTL genotype")
plotGenetpattern(Y=Y, cross=listeria, chr=chr,
                 main="by individual QTL genotype")
```

In both plots, we see one set of traits for which the QTL is dominant,
with the B allele associated with larger average trait value ($d
\approx -a$ and $a < 0$), and another set of traits for which the QTL
alleles are additive ($d \approx 0$), with the R allele associated
with larger average trait value ($a > 0$). This is evidence for two
QTL in the region, with different inheritance patterns.


### plot signed LOD score: `plotLODsign`

The function `plotLODsign` gives a second exploratory plot to
display the direction of the QTL
effects.  The input arguments are
`cross`, `Y`, and `chr`, as well as `addcovar` and `intcovar`, all as above.

For each trait, the estimated additive QTL effect is used as the sign of the
LOD score. We first run single-trait mapping to obtain LOD scores and
estimated QTL
positions. We use the R/qtl function `argmax.geno` to obtain imputed
QTL genotypes, and estimated the additive QTL effect as $\mu_{RR} - \mu_{BB}$.
Only traits with LOD
score bigger than a threshold will be displayed, the default value of
the threshold (`LOD.threshold`) is 3.

```{r,fig.cap="plot of signed LOD score for all the simulated triats"}
plotLODsign(Y=Y, cross=listeria, chr=chr)
```

We see that the traits mapping to the left side of the region all show
positive QTL effects, with the R allele associated with larger average
trait, while the traits mapping to the right side all show negative
QTL effects, with the B allele associated with larger average
trait. This is again evidence for two QTL in the region.

Tick marks at the bottom of the plot indicate the positions of genotyped markers.
We can pass these values to parameter `map`, but the default is pull this information from the
input `cross` object on the given chromosome, `chr`.

If we had already performed QTL analysis on the traits,
we can pass the signed LOD scores and their
mapped positions to `LODsign` and `maxPOS`, respectively.
By skipping the QTL analysis of the traits, this can speed up the
procedure. The use of
this feature will be shown at the last section.

### Linear discriminant plot: `plottrans.LDA`

In our final exploratory plot, we identify individuals with no
recombination event in the region of the QTL. For these individuals,
we will know their QTL genotype. We apply linear
discriminant analysis to the traits with the QTL genotype as the label
and plot the first two linear discriminants. If the QTL effect is
strong, this should show three distinct
clusters. We then calculate the linear discriminants for the
individuals that show a recombinantion event in the region, using the same
coffecients, and add them as points in the plot. If the recombinant individuals
fall within the clusters defined by the non-recombinants, this is
consistent with there being a single QTL. If, on the other hand, the
recombinants look distinctly different from the non-recombinants, this
suggests more than one QTL.

In the following code, we first find which individuals have no
recombinant event, and then call the function `plottrans.LDA` with
phenotype matrix `Y`, QTL genotype `qtlgeno` and nonrecombinant IDs
`nonrecomb`.

```{r, fig.cap="LDA plot for the simulated triats, non-recombinant mice are colored as yellow"}
out <- out[out$chr==chr, ]
m <- which(out$pos >= qtlpos-5 & out$pos <= qtlpos+5)
g <- apply(listeria$geno[[chr]]$prob[,m,], 1:2, which.max)
nonrecomb <- which(sapply(apply(g, 1, unique), length) == 1)
names(nonrecomb) <- rownames(Y)[nonrecomb]
plottrans.LDA(Y, qtlgeno, nonrecomb)
```

In the above plot, each point is a individual colored by its QTL
genotype if it is non-recombinant, or yellow if it was recombinant.
This particular example is not terribly informative about whether
there is one or two QTL.

## Analysis of a trans-eQTL hotspot with gene positions

Lastly, we will demonstrate the entire procedure, as recommended for
the analysis of
expression data. We start by mapping each phenotype individually,
then count the number of eQTL in windows of 10~cM.  When there seems
to be a trans-eQTL hotspot, we can build an object `transband` with
the function
`make.transbands`. This object contains basic information on the
transband: chromosome and position of the transband, number of eQTL in
the transband, and an estimate of the QTL position and LOD score under
the multivariate normal assumption. It also contains all of the information needed
for the exploratory plots: LOD scores and estimated QTL positions,
estimated additive and
dominance effects, QTL genotypes, and the IDs for the non-recombinant individuals.
By extracting such information from this object, we can run all of the
statistical tests and create the exploratory plots as a batch.

```{r, all, fig.width=12, fig.height=8, fig.cap="Summary plot for a trans-eQTL hotspot using simulated data"}
data(fake.probepos)
phenoname <- colnames(Y)
listeria$pheno <- data.frame(listeria$pheno, Y)
out <- scanone(listeria, pheno.col=phenoname, chr=1:19)
out1 <- convert_scan1(out, phenoname, chr=1:19)

marker.info <- get.marker.info(listeria, chr)
out.count <- count.trans(out1, fake.probepos, chr, marker.info)

par(mfrow=c(2,3))
plot(out.count, main="Count of trans-eQTL")

trans <- make.transbands(out1, fake.probepos, cross=listeria, chr=1:19,
                         mlratio = Y, lod.thr = 5,
                         trans.cM = 5, kernal.width = 1,
                         window.cM = 10, trans.count.thr = 0,
                         regn.cM = 5)
transband <- trans[[1]]

geno <- attr(transband, "geno")
nonrecomb <- attr(transband, "nonrecomb")
out <- attr(transband, "out")
map <- pull.map(listeria)

plot(obj, main="LOD profile")
plottrace(obj, main="Trace")
plotLODsign(maxPOS=out$pos, LODsign=sign(out$eff.a)*out$lod1, map=map[[chr]],
            main="Signed LOD")
plotGenetpattern(a=out$eff.a, d=out$eff.d, main="Inheritance Pattern")
plottrans.LDA(Y, geno, nonrecomb, main="LDA plot")

```

The results are as before, with one added plot in the top left: a
count of traits with a trans-eQTL in a sliding 10 cM window.


## Session Info

The following indicates the set of R packages that were used, as well
as the versions of R and the packages that we used.

```{r}
sessionInfo()
```
