<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Introduction to qtlpvl}
-->

```{r, echo = FALSE, message = FALSE}
library(qtlpvl)
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE)
```

# Introduction to qtlpvl

## Data 

We use data `hyper` from `R\qtl` as an example. A set of 10 phenotypes
are generated with two QTLs from chromosome 1, each affecting 5
phenotypes.

```{r}
set.seed(123456)
data(hyper)
hyper <- calc.genoprob(hyper)
geno <- pull.geno(hyper, chr="1")
genotype1 <- geno[,6]
genotype2 <- geno[,12]
n <- length(genotype1)
p <- 10
p1 <- floor(p/2)
G1 <- matrix(genotype1, n, p1)
G2 <- -matrix(genotype2, n, p-p1)
G <- cbind(G1, G2)
Y <- matrix(rnorm(n*p, sd=0.5), n, p)
Y <- Y + G * 0.2
```

## joint mapping: `scanone.mvn`

First we run a joint mapping with `scanone.mvn`, which assumes
multivariate normal model. 

```{r}
out <- scanone.mvn(hyper, Y, chr="1")
summary(out)
plot(out)
```

## test of 1 vs 2 QTL: `testpleio.1vs2`

We then test the hypothesis of $H_0$: There is only one QTL affecting
all the phenotypes; $H_1$: There are two QTLs, each affecting one of
the two phenotype groups. 

```{r}
obj <- testpleio.1vs2(cross=hyper, Y=Y, chr="1", n.simu=100,
                      region.l=19, region.r=90)
summary(obj)
plot(obj)
```
* red line: $L_1(\lambda_1) = \max_{\lambda_2} (\lambda_1, \lambda_2)$
* blue line: $L_2(\lambda_2) = \max_{\lambda_1} (\lambda_1, \lambda_2)$

```{r}
plottrace(obj)
```


## test of 1 vs p QTL: `testpleio.1vsp`
We could also test the hypothesis of $H_0$: There is only one QTL affecting
all the phenotypes; $H_1$: There are p QTLs, each affecting one of
the phenotypes. i.e., each pheotype is allowed to has its own QTL.

```{r}
obj2 <- testpleio.1vsp(cross=hyper, Y=Y, chr="1", n.simu=100)
summary(obj2)
plot(obj2)
```


## plot Genetic Pattern: `plotGenetpattern`

To illustrate the use of explorational plots, we generate data with
two different QTLs, the first QTL is additive, the second QTL is
dominant. We can tell from the plot the there are two different
genetic patterns, thus there might be two different QTLs, even if the
are closely linked.

```{r}
data(listeria)
geno <- pull.geno(listeria, chr="1")
genotype1 <- geno[,7]
genotype2 <- geno[,10]
n <- length(genotype1)
p <- 100
p1 <- floor(p/2)
G1 <- matrix(genotype1, n, p1)
G2 <- matrix(genotype2, n, p-p1)
G2[G2==3] <- 2
G2 <- -G2
G <- cbind(G1, G2)
Y <- matrix(rnorm(n*p,sd=0.5),n,p)
Y <- Y + G
plotGenetpattern(Y, genotype1)
```

## plot LOD score with sign: `plotLODsign` 
The direction of the QTL might be usfull as well.

```{r}
data(listeria)
listeria <- calc.genoprob(listeria)
geno <- pull.geno(listeria, chr="1")
genotype1 <- geno[,7]
genotype2 <- geno[,10]
n <- length(genotype1)
p <- 100
p1 <- floor(p/2)
G1 <- matrix(genotype1, n, p1)
G2 <- -matrix(genotype2, n, p-p1)
G <- cbind(G1, G2)
Y <- matrix(rnorm(n*p,sd=0.5),n,p)
Y <- Y + G
plotLODsign(listeria, Y, chr="1")
```


